Cybersecurity Impacts of the Covid-19 Pandemic in Italy

### 2021-03-31 article.tex
Minor fix
Added info of attacks wtr geographic area

\begin{enumerate}
\small
	\item \emph{Technical measures}
	\begin{itemize}
		\item Overall architecture of the digital security measures, integrated with the entire IT system architecture
		\item Physical countermeasures
		\item Identification, Authentication, and Authorization (IAA) measures
		\item Cybersecurity measures at local and geographical network level
		\item Measures for logical protection of each single Information and Communication Technology (ICT) system
		\item Application and software protection
		\item Data protection
	\end{itemize}
	\item \emph{Organizational measures}
	\begin{itemize}
		\item Organizational structure, roles, skills, certifications
		\item Organizational policies and procedures
		\item Contracts, agreements and digital security clauses with third parties
		\item Awareness and culture of digital security at all levels
		\item Auditing
	\end{itemize}
	\item \emph{Managerial and governmental measures}
	\begin{itemize}
		\item Digital security control, monitoring and management systems
		\item Disaster Recovery (DR) plan.
	\end{itemize}
\end{enumerate}

@@

The OAD 2020 survey
\footnote{The report is written in Italian,
  only the Executive Summary is in English: 186 A4 pages, 148 images and graphics, 11 Chapters
  (147 A4 pages) and 9 Attachments (39 A4 pages).}~\cite{oad20}
took place during the
Covid-19 pandemic and covered the entire 2019 and the first quarter
of 2020, when the pandemic exploded in Italy. The comparison of data from 2019 to that
from the first quarter of 2020, provided by the same set of respondents,
has statistical relevance and provides a 
clear indication of the Covid-19 pandemic on the intentional cyberattacks in Italy.
This pandemic has triggered a wide range of cyberattacks, mainly 
caused by the sudden, and partly unexpected, passage to working from home
and to a strong use of every type of IT service on the Internet, as a consequence
of the mobility lockdowns imposed by the Italian authorities. The resulting respondents pool
covers all the production sectors (Fig.~\ref{fig:9}), including public administration,
even if the majority of the respondents companies belong to the ICT sector (30.8\%).
However, many of them are represented by low percentages. this reason, not having sufficiently
representative samples, we have not considered particular data analysis and correlations by sectors.
The 2020 pool is well balanced for the size of the organizations, in terms
of employees, between those below 250 and the largest ones. It must be taken into account,
as reported by ISTAT~\cite{istat21}, that 99.91\%
of enterprises in Italy are SMEs, with up to 250 employees, and 95\% have fewer than ten employees.
The OAD 2020 survey is therefore able to consider small and 
tiny organizations, that are the vast majority in Italy, and that usually
are not considered in other cybersecurity surveys: 57.5\% of the OAD 2020 respondents
belong to structures with less than 250 employees and, of these, 22.1\% have fewer than ten
employees. The distribution of annual respondents' turnover also reflects the turnover of SMEs, as shown in Fig.~\ref{fig:7}. Regarding geographical distribution (Fig.~\ref{fig:8}), the responders' companies and organizations are located for the most part only on Italian territory (77.9\%), predominantly in northern Italy (43.3\%).

@@

\item In general, cybercriminals do not waste time and resources to attack small businesses,
  where they can only obtain a marginal profit and still
  take the risk of being caught by the police.  As confirmed by the trends shown
  in Fig.~\ref{fig:3}, attackers target the larger companies most,
  because the illegal gain is more interesting in that case.
\end{itemize}

\begin{figure}
  \centering
  \includegraphics[width=1\textwidth]{pictures/fig2.png}
  \caption{Trend of the reported attacks from the beginning of the OAD survey
    in 2008~\cite{oad20}. This comparison is only valid as a trend and not as a statistical
    measure, since the respondents are different from year to year. The dotted red line shows
    successive waves: after a relative peak of attacks, companies react and strengthen
    their cybersecurity measures, so that, the following year, the trend decreases.}
  \label{fig:2}
\end{figure}

@@

The impact of the pandemic on the cybersecurity in Italy is also highlighted by the data
provided by the Italian Postal and Telecommunication Police, that shows the situation for the
critical infrastructures (Table~\ref{tab:table1}), for the financial cyber-crimes
(Table~\ref{tab:table2}) and for the cyber-terrorism (Table~\ref{tab:table3}). 
Fraudulent transactions in 2019 were blocked for a total value of \euro 21.3 million and \euro 18 million were recovered; in the first quarter of 
2020, fraudulent transactions for \euro 20.2 million were blocked,
practically reaching in four months the amount
blocked in the full twelve months of 2019, and \euro 8 million have been
recovered. This is an indication of the increase in attacks on financial transactions
and services due to the very large use of these online services, caused by the mobility reduction
measures consequent of the Covid-19 pandemic.

\begin{table}[h]
  \begin{center}
  \resizebox{1\textwidth}{!}{
	\begin{tabular}{|r|r|r|r|r|r|}									
		\hline
		\textbf{Critical structure protection}	&	1 Jan - 30 Apr	&	1 Jan - 31 Dec	&	1 Jan - 31 Dec &	1 Jan - 31 Dec &	1 Jan - 31 Dec\\
                	&	2020	&	2019	&	2018 &	2017 &	2016\\
		\hline									
		Relevant attacks	&	282	&	1.181	&	459	&	1.032 & 844	\\
		Issued alerts	&	24.824	&	82.484	&	80.777	&	31.524 & 6.721	\\
		Initiated investigations 	&	34	&	155	&	74	&	72 & 70	\\
		People arrested	&	0	&	3	&	1	&	3	& 3\\
		People reported	&	0	&	117	&	14	&	1.316 &	1.226	\\
		Perquisitions	&	n.a.	&	n.a.	&	n.a.	&	73	& 58 \\
		\hline	
		Initiated investigations	&	12,05\%	&	13,12\%	&	16,12\%	&	6,98\%	& 8,29\% \\
		\hline	
		People arrested out of reported &	0\%	&	2,5\%	&	7,14\%	&	0,23\%	&	0,24\%\\						
		\hline	
	\end{tabular}
	}
    \end{center}

@@

\begin{table}[h]
  \begin{center}
  \resizebox{1\textwidth}{!}{
	\begin{tabular}{|r|r|r|r|r|r|}									
		\hline									
		\textbf{Financial Cyber Crime}	&	1 Jan - 30 Apr	&	1 Jan - 31 Dec	&	1 Jan - 31 Dec &	1 Jan - 31 Dec &	1 Jan - 31 Dec\\
                	&	2020	&	2019	&	2018 &	2017 &	2016\\
		\hline									
		Blocked fraudulent transactions	&	\textbf{\euro 20.200.000,00}	&	\textbf{\euro 21.333.990,00}	&	\euro 38.400.000,00	& \euro 20.839.576,00 & \euro 16.050.812,50 	\\
		Recovered amounts 	&	\euro 8.700.000,00	&	\euro 18.000.000,00	&	\euro 9.000.000,00	&	\euro 862.000,00 & n.a.	\\
		People arrested	&	n.a.	&	n.a.	&	n.a.	&	25	& 25\\
		People reported	&	n.a.	&	n.a.	&	n.a.	&	2.851 & 3.772	\\
		\hline	
		Recovered amounts out of fraud	&	12,05\%	&	13,12\%	&	16,12\%	&	6,98\%	& 8,29\% \\
		\hline	
	\end{tabular}
	}
    \end{center}
	\caption{Data of critical structure protection collected by Italian Postal and Telecommunication Police~\cite{oad20}.}									
	\label{tab:table2}									
\end{table}

\begin{table}[h]
  \begin{center}
	\begin{tabular}{|r|r|r|r|}									
		\hline									
		\textbf{Cyber-terrorism}	&	1 Jan - 30 Apr 2020	&	1 Jan - 31 Dec 2019	&	1 Jan - 31 Dec 2018\\
		\hline									
		Monitored web space	&	\euro 11.962	&	\euro 36.377	&	\euro 36.000\\
		Removed contents 	&	n.a.\	&	n.a.\	&	250\\
		\hline	
	\end{tabular}
        \end{center}
	\caption{Data of cyber terrorism collected by Italian Postal and Telecommunication Police~\cite{oad20}.}									
	\label{tab:table3}									
\end{table}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.90\textwidth]{pictures/fig6.png}
  \caption{Average diffusion (in percent) of the attack techniques detected in the most serious attacks between 2019 and the first quarter of 2020 (multiple answers)~\cite{oad20}.}
  \label{fig:6}
\end{figure}

>>>>

\begin{enumerate}
\tiny
	\item \emph{Technical measures}
	\begin{itemize}
		\item Overall architecture of the digital security measures, integrated with the entire IT system architecture
		\item Physical countermeasures
		\item Identification, Authentication, and Authorization (IAA) measures
		\item Cybersecurity measures at local and geographical network level
		\item Measures for logical protection of each single Information and Communication Technology (ICT) system
		\item Application and software protection
		\item Data protection
	\end{itemize}
	\item \emph{Organizational measures}
	\begin{itemize}
		\item Organizational structure, roles, skills, certifications
		\item Organizational policies and procedures
		\item Contracts, agreements and digital security clauses with third parties
		\item Awareness and culture of digital security at all levels
		\item Auditing
	\end{itemize}
	\item \emph{Managerial and governmental measures}
	\begin{itemize}
		\item Digital security control, monitoring and management systems
		\item Disaster Recovery (DR) plan.
	\end{itemize}
\end{enumerate}

@@

The OAD 2020 survey
\footnote{The report is written in Italian,
  only the Executive Summary is in English: 186 A4 pages, 148 images and graphics, 11 Chapters
  (147 A4 pages) and 9 Attachments (39 A4 pages).}~\cite{oad20}
took place during the
Covid-19 pandemic and covered the entire 2019 and the first quarter
of 2020, when the pandemic exploded in Italy. The comparison of data from 2019 to that
from the first quarter of 2020, provided by the same set of respondents,
has statistical relevance and provides a 
clear indication of the Covid-19 pandemic on the intentional cyberattacks in Italy.
This pandemic has triggered a wide range of cyberattacks, mainly 
caused by the sudden, and partly unexpected, passage to working from home
and to a strong use of every type of IT service on the Internet, as a consequence
of the mobility lockdowns imposed by the Italian authorities. The resulting respondents pool
covers all the production sectors (Fig.~\ref{fig:9}), including public administration,
even if the majority of the respondents companies belong to the ICT sector (30.8\%).
However, many of them are represented by low percentages. For this reason, not having sufficiently
representative samples, we have not considered particular data analysis and correlations by sectors.
The 2020 pool is well balanced for the size of the organizations, in terms
of employees, between those below 250 and the largest ones. It must be taken into account,
as reported by ISTAT~\cite{istat21}, that 99.91\%
of enterprises in Italy are SMEs, with up to 250 employees, and 95\% have fewer than ten employees.
The OAD 2020 survey is therefore able to consider small and 
tiny organizations, that are the vast majority in Italy, and that usually
are not considered in other cybersecurity surveys: 57.5\% of the OAD 2020 respondents
belong to structures with less than 250 employees and, of these, 22.1\% have fewer than ten
employees. The distribution of annual respondents' turnover also reflects the turnover of SMEs, as shown in Fig.~\ref{fig:7}. Regarding geographical distribution (Fig.~\ref{fig:8}), the responders' companies and organizations are located for the most part only on Italian territory (77.9\%), predominantly in northern Italy (43.3\%).

@@

\item In general, cybercriminals do not waste time and resources to attack small businesses,
  where they can only obtain a marginal profit and still
  take the risk of being caught by the police.  As confirmed by the trends shown
  in Fig.~\ref{fig:3} and Fig.~\ref{fig:4}, attackers target the larger companies most,
  because the illegal gain is more interesting in that case.
\end{itemize}

\begin{figure}
  \centering
  \includegraphics[width=0.9\textwidth]{pictures/fig2.png}
  \caption{Trend of the reported attacks from the beginning of the OAD survey
    in 2008~\cite{oad20}. This comparison is only valid as a trend and not as a statistical
    measure, since the respondents are different from year to year. The dotted red line shows
    successive waves: after a relative peak of attacks, companies react and strengthen
    their cybersecurity measures, so that, the following year, the trend decreases.}
  \label{fig:2}
\end{figure}

@@

\begin{figure}
	\begin{center}
		\includegraphics[scale=0.46]{pictures/fig4.png}
	\end{center}
  \caption{Trend of the reported attacks wrt. to the organizations' turnover in 2019.}
  \label{fig:4}
\end{figure}

Regarding the geographical distribution of attacks, Table.~\ref{tab:table4} shows
that occasional and infrequent attacks are evenly distributed throughout the territory in 2019,
with systematic attacks focused on northern Italy (where the companies with the highest turnover are present). 
While, Table.~\ref{tab:table5} shows that already in the first months of 2020 there were several attacks.

\begin{table}[ht]
  \begin{center}
  \resizebox{1\textwidth}{!}{
	\begin{tabular}{|l|r|r|r|}									
		\hline
		\textbf{Geographic area} &	Occasional cyber attacks	&	Frequent and repeated cyber attacks & Never detected cyber attacks \\
		& (few dozen in the year)	& (hundreds or more in the year) & \\
		\hline									
		Northern Italy	&	76,3\%	&	18,0\%	&	5,7\% \\
		Central Italy	&	46,5\%	&	0,0\%	&	53,5\%	\\
		Southern Italy and islands	&	80,8\%	&	0,0\% &	19,2\% \\
		Throughout Italy &	100,0\%	&	0,0\% &	0,0\% \\
		Throughout Italy and Europe &	58,8\%	&	0,0\%	&	41,2\%	\\
		Throughout Italy and abroad	(also outside Europe) &	100,0\%	&	0,0\%	&	0,0\% \\	
		\hline	
	\end{tabular}
  }
    \end{center}

	\caption{Distribution of digital attacks (detected by the respondents) wrt. to geographic area in 2019. The percentages of central Italy are 
	very different from the others because the organizations responding to that geographical area were numerically lower than the rest 
	of the survey participant (Fig.\ref{fig:8}).}									
	\label{tab:table4}									
\end{table}

\begin{table}[ht]
  \begin{center}
  \resizebox{1\textwidth}{!}{
	\begin{tabular}{|l|r|r|r|}									
		\hline
		\textbf{Geographic area} &	Occasional cyber attacks	&	Frequent and repeated cyber attacks & Never detected cyber attacks \\
		& (few dozen)	&  (hundreds or more) & \\
		\hline									
		Northern Italy	&	5,8\%	&	6,3\%	&	87,9\% \\
		Central Italy	&	0,0\%	&	0,0\%	&	100,0\%	\\
		Southern Italy and islands	&	0,0\%	&	0,0\% &	100,0\% \\
		Throughout Italy &	30,3\%	&	0,0\%	&	69,7\%	\\
		Throughout Italy and Europe &	0,0\%	&	0,0\%	&	100,0\%	\\
		Throughout Italy and abroad (also outside Europe)	&	28,8\%	&	0,0\%	&	71,2\%	\\					
		\hline		
	\end{tabular}
  }
    \end{center}

	\caption{Distribution of digital attacks (detected by the respondents) wrt. to geographic area in 1Q-2020.}									
	\label{tab:table5}									
\end{table}

The impact of the pandemic on the cybersecurity in Italy is also highlighted by the data
provided by the Italian Postal and Telecommunication Police, that shows the situation for the
critical infrastructures (Table~\ref{tab:table1}), for the financial cyber-crimes
(Table~\ref{tab:table2}) and for the cyber-terrorism (Table~\ref{tab:table3}). 
Fraudulent transactions all Italyin 2019 were blocked for a total value of \euro 21.3 million and \euro 18 million were recovered; in the first quarter of 
2020, fraudulent transactions for \euro 20.2 million were blocked,
practically reaching in four months the amount
blocked in the full twelve months of 2019, and \euro 8 million have been
recovered. This is an indication of the increase in attacks on financial transactions
and services due to the very large use of these online services, caused by the mobility reduction
measures consequent of the Covid-19 pandemic.

\begin{table}[ht]
  \begin{center}
  \resizebox{1\textwidth}{!}{
	\begin{tabular}{|r|r|r|r|r|r|}									
		\hline
		\textbf{Critical structure protection}	&	1 Jan - 30 Apr	&	1 Jan - 31 Dec	&	1 Jan - 31 Dec &	1 Jan - 31 Dec &	1 Jan - 31 Dec\\
                	&	2020	&	2019	&	2018 &	2017 &	2016\\
		\hline									
		Relevant attacks	&	282	&	1.181	&	459	&	1.032 & 844	\\
		Issued alerts	&	24.824	&	82.484	&	80.777	&	31.524 & 6.721	\\
		Initiated investigations 	&	34	&	155	&	74	&	72 & 70	\\
		People arrested	&	0	&	3	&	1	&	3	& 3\\
		People reported	&	0	&	117	&	14	&	1.316 &	1.226	\\
		Perquisitions	&	n.a.	&	n.a.	&	n.a.	&	73	& 58 \\
		\hline	
		Initiated investigations	&	12,05\%	&	13,12\%	&	16,12\%	&	6,98\%	& 8,29\% \\
		\hline	
		People arrested out of reported &	0\%	&	2,5\%	&	7,14\%	&	0,23\%	&	0,24\%\\						
		\hline	
	\end{tabular}
	}
    \end{center}

@@

\begin{table}[ht]
  \begin{center}
  \resizebox{1\textwidth}{!}{
	\begin{tabular}{|r|r|r|r|r|r|}									
		\hline									
		\textbf{Financial Cyber Crime}	&	1 Jan - 30 Apr	&	1 Jan - 31 Dec	&	1 Jan - 31 Dec &	1 Jan - 31 Dec &	1 Jan - 31 Dec\\
                	&	2020	&	2019	&	2018 &	2017 &	2016\\
		\hline									
		Blocked fraudulent transactions	&	\textbf{\euro 20.200.000,00}	&	\textbf{\euro 21.333.990,00}	&	\euro 38.400.000,00	& \euro 20.839.576,00 & \euro 16.050.812,50 	\\
		Recovered amounts 	&	\euro 8.700.000,00	&	\euro 18.000.000,00	&	\euro 9.000.000,00	&	\euro 862.000,00 & n.a.	\\
		People arrested	&	n.a.	&	n.a.	&	n.a.	&	25	& 25\\
		People reported	&	n.a.	&	n.a.	&	n.a.	&	2.851 & 3.772	\\
		\hline	
		Recovered amounts out of fraud	&	12,05\%	&	13,12\%	&	16,12\%	&	6,98\%	& 8,29\% \\
		\hline	
	\end{tabular}
	}
    \end{center}
	\caption{Data of critical structure protection collected by Italian Postal and Telecommunication Police~\cite{oad20}.}									
	\label{tab:table2}									
\end{table}

\begin{table}[ht]
  \begin{center}
	\begin{tabular}{|r|r|r|r|}									
		\hline									
		\textbf{Cyber-terrorism}	&	1 Jan - 30 Apr 2020	&	1 Jan - 31 Dec 2019	&	1 Jan - 31 Dec 2018\\
		\hline									
		Monitored web space	&	\euro 11.962	&	\euro 36.377	&	\euro 36.000\\
		Removed contents 	&	n.a.\	&	n.a.\	&	250\\
		\hline	
	\end{tabular}
        \end{center}
	\caption{Data of cyber terrorism collected by Italian Postal and Telecommunication Police~\cite{oad20}.}									
	\label{tab:table3}									
\end{table}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.85\textwidth]{pictures/fig6.png}
  \caption{Average diffusion (in percent) of the attack techniques detected in the most serious attacks between 2019 and the first quarter of 2020 (multiple answers)~\cite{oad20}.}
  \label{fig:6}
\end{figure}
